An elaborate data set on human gait and the effect of mechanical perturbations

### 2015-02-19 paper.tex

\section*{Introduction}
%
The collection of dynamical data during human walking has a long history
beginning with the first motion pictures and now with modern marker based
motion capture techniques and high fidelity ground reaction load measurements.
Even though years of data on thousands of subjects now exist, this data is not
widely disseminated, well organized, nor available with few or no restrictions.
David Winter's published normative gait data, \cite{Winter1990}, is widely used
in biomechanical studies, yet it comes from few subjects and only a small
number of gait cycles per subject. This small source has successfully enabled
many other studies, such as powered prosthetic control design, \cite{Sup2008},
but success in other research fields using large sets of data for discovery
lead one to believe that more elaborate data sets may benefit the field of
human motion studies. To enable such work, biomechanical data needs to be
shared extensively, organized, and curated to enable future analysts.

@@

\cite{Chester2007} report on a large gait database comparison where one
database contained kinematic data of 409 gait cycles of children from 1 to 7
years old but the data does not seem to be publicly available. This is,
unfortunately, typical. Tirosh et. al, recognized the need for a comprehensive
data base for clinical gait data and created the Gaitabase, \cite{Tirosh2010}.
This database may contain a substantial amount of data but it is encumbered by
a complicated and restrictive license and sharing scheme. However, there
are examples of data with less restrictions. The University of Wisconsin at
LaCrosse has an easily accessible normative gait data set, \cite{Willson2014},
from 25 subjects with lower extremity marker data from multiple gait cycles and
force plate measurements from a single gait cycle.

@@

With all of this in mind, we collected over seven and half hours of gait data
from fifteen able bodied subjects which amounts to over 25,000 gait cycles. The
subjects walked at three different speeds on an instrumented treadmill while we
collected full body marker locations and ground reaction loads from a pair of
force plates. The protocol for the majority of the trials included two minutes
of normal walking and eight minutes of walking under the influence of
pseudo-random belt speed fluctuations. The data has been organized complete
with rich meta data and made available in the most unrestrictive form for other
research uses following modern best practices in data sharing, \cite{White2013}.

@@

When belt speed is not constant, the inertia of the rollers and motor will
induce error in the force plate $x$ axis moment, and hence, the
anterior-posterior coordinate ($z$ axis) of the center of pressure that is
measured by the instrumentation in the treadmill. If needed, this error can be
compensated by a linear model as shown by \cite{Hnat2014}. The model
coefficients can be identified from the unloaded trials given in
Table~\ref{tab:comptrials}. The error due to inertia is random and does not
affect the averaged joint moments presented in
Figure~\ref{fig:angle-torque-comparison}. Compensation should, however, be done
if joint moments from individual gait cycles are of interest rather than the
ensemble average.
%
\begin{table}
  \cprotect\caption{A list of unloaded trials collected for each speed. Each
    loaded trial includes a compensation file listed in its meta data which
    matches it to these unloaded trials. Generated by
    \verb|src/subject_table.py|.}
  \centering
  \small
  \input{tables/compensation_trials.tex}
  \label{tab:comptrials}
\end{table}

@@

The \verb+trial+ section contains the information about the particular trial.
Each trial has a unique identification number along with a variety of other
information, detailed below:
%
\begin{description}
  \item[analog-channel-map] A mapping of the strings D-Flow assigns to signals
    emitted from the analog channels of the NI USB-6255 to names the user
    desires.
  \item[cortex-version] The version of Cortex used to record the trial.
  \item[datetime] A date formatted string giving the date of the trial in the
    \verb|YYYY-MM-DD| format.
  \item[dflow-version] The version of D-Flow used to record the trial.
  \item[events] A key value map which prescribes names to the alphabetic events
    recorded in the record file.
  \item[files] A key value mapping of files associated with this trial
    where the key is the D-Flow file type and the value is the path to the file
    relative to the meta file. The compensation file corresponds to an unloaded
    trial collected on the same day that could be used for inertial
    compensation purposes, if needed.
  \item[hardware-settings] There are tons of settings for the hardware in both
    D-Flow, Cortex, and the other software in the system. This contains any
    non-default settings.
    \begin{description}
      \item[high-performance] A boolean value indicating whether the D-Flow high
        performance setting was on (True) or off (False).
    \end{description}
  \item[id] An unique three digit integer identifier for the trial. All of the
    file names and directories associated with this trial include this number.
  \item[marker-map] A key value map which maps marker names in the mocap file
    to the user's desired names for the markers.
  \item[marker-set] Indicates the HBM~\cite{Bogert2013} marker set used during
    the trial, either full, lower, or NA.
  \item[nominal-speed] A float representing the nominal desired treadmill speed
    during the trial.
  \item[nominal-speed-units] A string providing the units of the nominal speed.
  \item[notes] Any notes about the trial.
  \item[pitch] A boolean that indicates if the treadmill pitch degree of
    freedom was actuated during the trial.
  \item[stationary-platform] A boolean that indicates whether the treadmill
      sway or pitch motion was actuated during the trial. If this flag is
      false, the measured ground reaction loads must be compensated for the
      inertial affects and be expressed in the motion capture reference frame.
  \item[subject-id] An integer corresponding to the subject in the trial.
  \item[sway] A boolean that indicates if the treadmill lateral degree of
    freedom was actuated during the trial.
\end{description}

@@

\subsection*{Processed Data}
%
We developed a toolkit for data processing, GaitAnalysisToolKit v0.1.2,
\cite{Moore2014a}, for common gait computations and provide an example
processed trial to present the nature of the data. The tool was developed in
Python, is dependent on the SciPy Stack [NumPy~\citep{Walt2011},
SciPy~\citep{Jones2001}, matplotlib~\citep{Hunter2007},
Pandas~\citep{McKinney2010}, etc] and Octave~\citep{Octave2014}, and provides
two main classes: one to do basic gait data cleaning from D-Flow's output
files, \verb|DFlowData|, and a second to compute common gait variables of
interest, \verb|GaitData|.

@@

\section*{Data Availability}
%
The data set, \cite{Moore2014}, is available via the Zenodo data repository.
Two approximately 1.2GB gzipped tar balls contain the data and a \verb|README|
file with a short description of the contents. The data is released under the
Creative Commons CC0 license (http://creativecommons.org/about/cc0) following
best practices for sharing scientific data.

\section*{Software Availability}
%
The tables, figures, and the paper can be reproduced from the source
repository shared on Github: https://github.com/csu-hmc/perturbed-data-paper.
Along with the source code in the repository, the computations depend on
version 0.1.2 of the GaitAnalysisToolKit, \cite{Moore2014a}, which can be
downloaded from Zenodo or the Python Package Index (http://pypi.python.org).

>>>>

\section*{Introduction}
%
The collection of dynamical data during human walking has a long history
beginning with the first motion pictures and now with modern marker based
motion capture techniques and high fidelity ground reaction load measurements.
Even though years of data on thousands of subjects now exist, this data is not
widely disseminated, well organized, nor available with few or no restrictions.
David Winter's published normative gait data \citep{Winter1990} is widely used
in biomechanical studies, yet it comes from few subjects and only a small
number of gait cycles per subject. This small source has successfully enabled
many other studies, such as powered prosthetic control design \citep{Sup2008}
but success in other research fields using large sets of data for discovery
lead one to believe that more elaborate data sets may benefit the field of
human motion studies. To enable such work, biomechanical data needs to be
shared extensively, organized, and curated to enable future analysts.

@@

\cite{Chester2007} report on a large gait database comparison where one
database contained kinematic data of 409 gait cycles of children from 1 to 7
years old but the data does not seem to be publicly available. This is,
unfortunately, typical. \cite{Tirosh2010} recognized the need for a
comprehensive data base for clinical gait data and created the Gaitabase. This
database may contain a substantial amount of data but it is encumbered by a
complicated and restrictive license and sharing scheme. However, there are
examples of data with less restrictions. The University of Wisconsin at
LaCrosse has an easily accessible normative gait data set \citep{Willson2014}
from 25 subjects with lower extremity marker data from multiple gait cycles and
force plate measurements from a single gait cycle.

@@

With all of this in mind, we collected over seven and half hours of gait data
from fifteen able bodied subjects which amounts to over 25,000 gait cycles. The
subjects walked at three different speeds on an instrumented treadmill while we
collected full body marker locations and ground reaction loads from a pair of
force plates. The protocol for the majority of the trials included two minutes
of normal walking and eight minutes of walking under the influence of
pseudo-random belt speed fluctuations. The data has been organized complete
with rich meta data and made available in the most unrestrictive form for other
research uses following modern best practices in data sharing \citep{White2013}.

@@

When belt speed is not constant, the inertia of the rollers and motor will
induce error in the force plate $x$ axis moment, and hence, the
anterior-posterior coordinate ($z$ axis) of the center of pressure that is
measured by the instrumentation in the treadmill. If needed, this error can be
compensated by a linear model as shown in \cite{Hnat2014}. The model
coefficients can be identified from the unloaded trials given in
Table~\ref{tab:comptrials}. The error due to inertia is random and does not
affect the averaged joint moments presented in
Figure~\ref{fig:angle-torque-comparison}. Compensation should, however, be done
if joint moments from individual gait cycles are of interest rather than the
ensemble average.
%
\begin{table}
  \cprotect\caption{A list of unloaded trials collected for each speed. Each
    loaded trial includes a compensation file listed in its meta data which
    matches it to these unloaded trials. Generated by
    \verb|src/subject_table.py|.}
  \centering
  \small
  \input{tables/compensation_trials.tex}
  \label{tab:comptrials}
\end{table}

@@

The \verb+trial+ section contains the information about the particular trial.
Each trial has a unique identification number along with a variety of other
information, detailed below:
%
\begin{description}
  \item[analog-channel-map] A mapping of the strings D-Flow assigns to signals
    emitted from the analog channels of the NI USB-6255 to names the user
    desires.
  \item[cortex-version] The version of Cortex used to record the trial.
  \item[datetime] A date formatted string giving the date of the trial in the
    \verb|YYYY-MM-DD| format.
  \item[dflow-version] The version of D-Flow used to record the trial.
  \item[events] A key value map which prescribes names to the alphabetic events
    recorded in the record file.
  \item[files] A key value mapping of files associated with this trial
    where the key is the D-Flow file type and the value is the path to the file
    relative to the meta file. The compensation file corresponds to an unloaded
    trial collected on the same day that could be used for inertial
    compensation purposes, if needed.
  \item[hardware-settings] There are tons of settings for the hardware in both
    D-Flow, Cortex, and the other software in the system. This contains any
    non-default settings.
    \begin{description}
      \item[high-performance] A boolean value indicating whether the D-Flow high
        performance setting was on (True) or off (False).
    \end{description}
  \item[id] An unique three digit integer identifier for the trial. All of the
    file names and directories associated with this trial include this number.
  \item[marker-map] A key value map which maps marker names in the mocap file
    to the user's desired names for the markers.
  \item[marker-set] Indicates the HBM~\citep{Bogert2013} marker set used during
    the trial, either full, lower, or NA.
  \item[nominal-speed] A float representing the nominal desired treadmill speed
    during the trial.
  \item[nominal-speed-units] A string providing the units of the nominal speed.
  \item[notes] Any notes about the trial.
  \item[pitch] A boolean that indicates if the treadmill pitch degree of
    freedom was actuated during the trial.
  \item[stationary-platform] A boolean that indicates whether the treadmill
      sway or pitch motion was actuated during the trial. If this flag is
      false, the measured ground reaction loads must be compensated for the
      inertial affects and be expressed in the motion capture reference frame.
  \item[subject-id] An integer corresponding to the subject in the trial.
  \item[sway] A boolean that indicates if the treadmill lateral degree of
    freedom was actuated during the trial.
\end{description}

@@

\subsection*{Processed Data}
%
We developed a toolkit for data processing, GaitAnalysisToolKit v0.1.2
\citep{Moore2014a} for common gait computations and provide an example
processed trial to present the nature of the data. The tool was developed in
Python, is dependent on the SciPy Stack [NumPy~\citep{Walt2011},
SciPy~\citep{Jones2001}, matplotlib~\citep{Hunter2007},
Pandas~\citep{McKinney2010}, etc] and Octave~\citep{Octave2014}, and provides
two main classes: one to do basic gait data cleaning from D-Flow's output
files, \verb|DFlowData|, and a second to compute common gait variables of
interest, \verb|GaitData|.

@@

\section*{Data Availability}
%
The data set \citep{Moore2014} is available via the Zenodo data repository.
Two approximately 1.2GB gzipped tar balls contain the data and a \verb|README|
file with a short description of the contents. The data is released under the
Creative Commons CC0 license (http://creativecommons.org/about/cc0) following
best practices for sharing scientific data.

\section*{Software Availability}
%
The tables, figures, and the paper can be reproduced from the source
repository shared on Github: https://github.com/csu-hmc/perturbed-data-paper.
Along with the source code in the repository, the computations depend on
version 0.1.2 of the GaitAnalysisToolKit \citep{Moore2014a} which can be
downloaded from Zenodo or the Python Package Index (http://pypi.python.org).
